---
title: "Estimated Demand"
author: "Sterling Fearing"
date: "4/25/2022"
output: html_document
---
Goal: Measure estimated demand for abortion services on the county level using proportions provided by public health annual review for age, race, and poverty. (https://www.annualreviews.org/doi/full/10.1146/annurev.publhealth.26.021304.144351)

Age: 
15-17: 15 per 1000
18-19: 39 per 1000
20-24: 47 per 1000
25-29: 32 per 1000
30-34: 17 per 1000
35-39: 9 per 1000
40 and up: 3 per 1000

Income:
44 per 1000 among low-income women (below 200% of FPL)
10 per 1000 among high income

Race:
13 per 1000 for White women
49 per 1000 for Black women
31 per 1000 for Asian/Pacific Islander women
33 per 1000 for Hispanic women



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# load libraries

library(sf)
library(tidycensus)
library(tidyverse)
library(tigris)
library(tidyr)
library(here)

```
```{r}
# load existing data sets

# race percent from Community Health Rankings
race <- read_csv(here("code", "data", "racePercent.csv"))

# cleaned census data from countycensus_data.Rmd
census <- read_csv(here("code", "data", "censusDataCounty.csv"))

```


```{r}
# Variables 

# Age 
## identify variables from B01001 : AGE & SEX table
#  1. Female  : Age 15-17  B01001_030 Estimate!!Total!!Female!!15 to 17 years
#  2. Female  : Age 18-19  B01001_031 Estimate!!Total!!Female!!18 and 19 years
#  3. Female  : Age 20     B01001_032 Estimate!!Total!!Female!!20 years
#  4. Female  : Age 21     B01001_033 Estimate!!Total!!Female!!21 years
#  5. Female  : Age 22-24  B01001_034 Estimate!!Total!!Female!!22 to 24 years
#  6. Female  : Age 25-29  B01001_035 Estimate!!Total!!Female!!25 to 29 years
#  7. Female  : Age 30-34  B01001_036 Estimate!!Total!!Female!!30 to 34 years
#  8. Female  : Age 35-39  B01001_037 Estimate!!Total!!Female!!35 to 39 years
#  9. Female  : Age 40-44  B01001_038 Estimate!!Total!!Female!!40 to 44 years
# 10. Female  : Age 45-49  B01001_039 Estimate!!Total!!Female!!45 to 49 years

# Race
## Source: County Health Rankings
# 1. Percent Black
# 2. Percent Asian
# 3. Percent Hispanic
# 4. Percent White 

# Income
# 1. Poverty : S1701_C01_042 Estimate!!Total below 200% poverty level

```

```{r}
# Get ACS data

ACSdata <- get_acs(geography = 'county', variables = c(age15_17 = "B01001_030",
                                                        age18_19 = "B01001_031",
                                                        age20 = "B01001_032",
                                                        age21 = "B01001_033",
                                                        age22_24 = "B01001_034",
                                                        age25_29 = "B01001_035",
                                                        age30_34 = "B01001_036",
                                                        age35_39 = "B01001_037",
                                                        age40_44 = "B01001_038",
                                                        age45_49 = "B01001_039",
                                                        popTot =  "B02001_001",
                                                        lowincome = "S1701_C01_042"),
                    year = 2018, geometry = FALSE)


```



```{r}
# Tidy new ACS data

ACSdata1 <- ACSdata %>%
  select(-moe) %>%
  pivot_wider(names_from = variable,
              values_from = estimate) %>%
  filter(!str_detect(NAME, "Hawaii|Alaska|Puerto Rico")) %>%
  mutate(age20_24 = age20 + age21 + age22_24,
         age40_49 = age40_44 + age45_49) #create columns that match age categories

```

```{r}
# select variables of interest from ACS data
ACSdata2 <- ACSdata1 %>%
  select(GEOID, age15_17, age18_19, age20_24, age25_29, age30_34, age35_39, age40_49, popTot, lowincome)

# select variables of interest from previous censusDataCounty table

census1 <- census %>%
  select(GEOID, age15_49, povPerc, pciE)
```

```{r}
# Join tables

joinDemand <- full_join(race, ACSdata2, by= "GEOID")
joinDemand2 <- full_join(joinDemand, census, by= "GEOID")

```

```{r}
# Estimate demand by age group
#15-17: 15 per 1000
#18-19: 39 per 1000
#20-24: 47 per 1000
#25-29: 32 per 1000
#30-34: 17 per 1000
#35-39: 9 per 1000
#40-49 (over 40): 4 per 1000

demandAGE <- joinDemand2 %>%
  mutate(age15_17D = (age15_17*15)/1000) %>%
  mutate(age18_19D = (age18_19*39)/1000) %>%
  mutate(age20_24D = (age20_24*47)/1000) %>%
  mutate(age25_29D = (age25_29*32)/1000) %>%
  mutate(age30_34D = (age30_34*17)/1000) %>%
  mutate(age35_39D = (age35_39*9)/1000) %>%
  mutate(age40_49D = (age40_49*4)/1000)

# add groups to determine demand based on age in each county

demandAGEadd <- demandAGE %>%
  mutate(age15_49D = age15_17D + age18_19D + age20_24D + age25_29D + age30_34D + age35_39D + age40_49D)


```

```{r}
# Estimate demand by race

demandRACE <- joinDemand2 %>%
  mutate(whiteD = age15_49*whiteR) %>%
  mutate(blackD = age15_49*blackR) %>%
  mutate(asianD = age15_49*asianR) %>%
  mutate(hispD = age15_49*hispR)

# add groups to determine demand based on race in each county

demandRACEadd <- demandRACE %>%
  mutate(raceD = whiteD + blackD + asianD + hispD)
```

```{r}
# Estimate demand by income 

demandINCOME <- 
```

