---
title: "Estimated Demand"
author: "Sterling Fearing"
date: "4/25/2022"
output: html_document
---
Goal: Measure estimated demand for abortion services on the county level using proportions provided by public health annual review and Guttmacher Instutute for age, race, and poverty. 

References: Patterns in the Socioeconomic Characteristics of Women Obtaining Abortions (2000), Guttmacher Institute
(See Table 1, column = Abortion Rate 2000)
Public Health Annual Review (https://www.annualreviews.org/doi/full/10.1146/annurev.publhealth.26.021304.144351)

Age: 
15-17: 15 per 1000
18-19: 39 per 1000
20-24: 47 per 1000
25-29: 32 per 1000
30-34: 17 per 1000
35-39: 9 per 1000
40 and up: 3 per 1000

Income: 
~ 44 per 1000 below 100% poverty level
~ 38 per 1000 between 100-199% poverty level
~ 21 per 1000 above 200% poverty level
note- census data does not provide data on pop of reproductive aged women above 300% of FPL

Race:
13 per 1000 for White women
49 per 1000 for Black women
31 per 1000 for Asian/Pacific Islander women
33 per 1000 for Hispanic women



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# load libraries
library(tidycensus)
library(tidyverse)
library(here)

```

```{r}
# load existing data sets

# race percent from Community Health Rankings
race <- read_csv(here("code", "data", "racePercent.csv"))

# cleaned census data from censusData.Rmd
ACS <- read_csv(here("code", "data", "ACSdata.csv")) #ACS version edited with Numbers to add est. pov for Rio Arriba, NM

```

```{r}
# select variables of interest from ACS data
ACSdata1 <- ACS %>%
  select(GEOID, age15_17, age18_19, age20_24, age25_29, age30_34, age35_39, age40_49, age15_49, povWtotal, povW100, povW100_199, povW200)

```

```{r}
# Join tables

joinD <- full_join(race, ACSdata1, by= "GEOID")

```

```{r}
# Estimate demand by age group
#15-17: 15 per 1000
#18-19: 39 per 1000
#20-24: 47 per 1000
#25-29: 32 per 1000
#30-34: 17 per 1000
#35-39: 9 per 1000
#40-49 (over 40): 4 per 1000

demandAGE <- joinD %>%
  mutate(age15_17D = (age15_17*15)/1000) %>%
  mutate(age18_19D = (age18_19*39)/1000) %>%
  mutate(age20_24D = (age20_24*47)/1000) %>%
  mutate(age25_29D = (age25_29*32)/1000) %>%
  mutate(age30_34D = (age30_34*17)/1000) %>%
  mutate(age35_39D = (age35_39*9)/1000) %>%
  mutate(age40_49D = (age40_49*4)/1000)

# add groups to determine total demand in county based on age 

demandAGEadd <- demandAGE %>%
  mutate(age15_49D = age15_17D + age18_19D + age20_24D + age25_29D + age30_34D + age35_39D + age40_49D)

# select variables of interest

demandAGE2 <- demandAGEadd %>%
  select(GEOID, age15_17D, age18_19D, age20_24D, age25_29D, age30_34D, age35_39D, age40_49D, age15_49D)

```

```{r}
# Estimate demand by race
#13 per 1000 for White women
#49 per 1000 for Black women
#31 per 1000 for Asian/Pacific Islander women
#33 per 1000 for Hispanic women

estRACE <- joinD %>%
  mutate(whiteE = age15_49*whiteR) %>%
  mutate(blackE = age15_49*blackR) %>%
  mutate(asianE = age15_49*asianR) %>%
  mutate(hispE = age15_49*hispR)

demandRACE <- estRACE %>%
  mutate(whiteD = (whiteE*13)/1000) %>%
  mutate(blackD = (blackE*49)/1000) %>%
  mutate(asianD = (asianE*31)/1000)%>%
  mutate(hispD = (hispE*33)/1000)
  

# add groups to determine demand based on race in each county

demandRACEadd <- demandRACE %>%
  mutate(raceD = whiteD + blackD + asianD + hispD)

# select variables of interest
demandRACE2 <- demandRACEadd %>%
  select(GEOID, whiteD, blackD, asianD, hispD, raceD)
```

```{r}
# Estimate demand by income 

#~ 44 per 1000 below 100% poverty level
#~ 38 per 1000 between 100-199% poverty level
#~ 21 per 1000 above 200% poverty level
#note- census data does not provide data on pop of reproductive aged women above 300% of FPL

demandINCOME <- joinD %>%
  mutate(p100D = (povW100*44)/1000) %>%
  mutate(p100_199D = (povW100_199*38)/1000) %>%
  mutate(p200D = (povW200*21)/1000)

demandINCOMEadd <- demandINCOME %>%
  mutate(incomeD = p100D + p100_199D + p200D)

demandINCOME2 <- demandINCOMEadd %>%
  select(GEOID, p100D, p100_199D, p200D, incomeD)
```

```{r}
# join age, race, income tables to create demand table

joinAR <- left_join(demandAGE2, demandRACE2, by= "GEOID") 

joinARI <- left_join(joinAR, demandINCOME2, by= "GEOID")
```


```{r}
# split GEOID to create stateFIPS and countyFIPS columns

splitGEO <- joinARI %>%
  separate(GEOID, into = c('stateFIPS', 'countyFIPS'), sep = 2, remove = FALSE)
```

```{r}
# save as CSV

write_csv(splitGEO, here("code/data/demandIndex.csv"))
```

