title: "State Abortion Law Index"
author: "Sterling Fearing"
date: "9/19/2021"
output: html_document
---

The data used for the Abortion Law Index is sourced from the Guttmacher Institute. The link to for online table can be found [here](https://www.guttmacher.org/state-policy/explore/overview-abortion-laws). This data was sourced on August 12, 2021. The table was copied into Microsoft excel and saved as a csv file. 

The gestational limit for Texas was changed to 6 weeks to reflect the new abortion law which took effect in September 2021. The bill bans abortions if an ultrasound detects what lawmakers have defined as a "fetal heartbeat", which can occur as early as 6 weeks LMP (last menstrual period). Medical experts argue that this term is misleading because a heart is not present in the embryo at this stage in development. 

An additional variable that identifies the states where in-person counseling laws necessitate two trips to the clinic was added to the Overview of State Abortion Law Table. This data can be found [here](https://www.guttmacher.org/state-policy/explore/counseling-and-waiting-periods-abortion). The special characters used in the column were converted in Microsoft Excel to avoid conflict with definitions provided in primary table. 
For reference:
ξ = X100 to represent the definition "In-person counseling is not required for patients who live more than 100 miles from an abortion provider."
Э = Xtele and Xany for Kentucky and Utah respectively to represent the definition "In Kentucky, a patient may be able to use telemedicine for abortion counseling. In Utah, a patient may obtain abortion counseling in person at any medical office in the state."


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load libraries
library(purrr)
library(tidyverse)
library(here)
```

```{r}
# Load data

lawRaw <- read_csv(here("code/data/laws.csv"))

view(lawRaw)

```
