---
title: "State Abortion Law Index"
author: "Sterling Fearing"
date: "9/19/2021"
output: html_document
---

The data used for the Abortion Law Index is sourced from the Guttmacher Institute. The link to for online table can be found [here](https://www.guttmacher.org/state-policy/explore/overview-abortion-laws). This data was sourced on August 12, 2021. The table was copied into Microsoft excel and saved as a csv file. 

The gestational limit for Texas was changed to 6 weeks to reflect the new abortion law which took effect in September 2021. The bill bans abortions if an ultrasound detects what lawmakers have defined as a "fetal heartbeat", which can occur as early as 6 weeks LMP (last menstrual period). Medical experts argue that this term is misleading because a heart is not present in the embryo at this stage in development. 

An additional variable that identifies the states where in-person counseling laws necessitate two trips to the clinic was added to the Overview of State Abortion Law Table. This data can be found [here](https://www.guttmacher.org/state-policy/explore/counseling-and-waiting-periods-abortion). The special characters used in the column were converted in Microsoft Excel to avoid conflict with definitions provided in primary table. 
For reference:
ξ = X100 to represent the definition "In-person counseling is not required for patients who live more than 100 miles from an abortion provider."
Э = Xtele and Xany for Kentucky and Utah respectively to represent the definition "In Kentucky, a patient may be able to use telemedicine for abortion counseling. In Utah, a patient may obtain abortion counseling in person at any medical office in the state."


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load libraries
library(purrr)
library(tidyverse)
library(here)
```

```{r}
# Load data

lawRaw <- read_csv(here("code/data/laws.csv"))

view(lawRaw)

```

```{r}
# Clean data

# select variables of interest
lawSelect <- lawRaw %>%
   select("STATE", "SECOND PHYSICIAN MUST PARTICIPATE IF AT:", "PROHIBITED EXCEPT IN CASES OF LIFE OR HEALTH ENDANGERMENT IF AT:", "\"PARTIAL-BIRTH\" ABORTION BANNED", "Funds All or Most Medically Necessary Abortions", "Funds Limited  to Life Endangerment, Rape and Incest", "PRIVATE INSURANCE COVERAGE LIMITED", "PARENTAL INVOLVEMENT REQUIRED FOR MINORS", "In-Person Counseling Necessitates Two Trips to Clinic")

# rename variables
lawRename <- lawSelect %>%
   select(state = `STATE`,
          twoPhys = `SECOND PHYSICIAN MUST PARTICIPATE IF AT:`,
          gestLim = `PROHIBITED EXCEPT IN CASES OF LIFE OR HEALTH ENDANGERMENT IF AT:`,
          pbBan = `\"PARTIAL-BIRTH\" ABORTION BANNED`,
          pubFundAll = `Funds All or Most Medically Necessary Abortions`,
          pubFundLim = `Funds Limited  to Life Endangerment, Rape and Incest`,
          privLim = `PRIVATE INSURANCE COVERAGE LIMITED`,
          parent = `PARENTAL INVOLVEMENT REQUIRED FOR MINORS`,
          twoTrips = `In-Person Counseling Necessitates Two Trips to Clinic`)

```

```{r}

# Create new columns with index values
   
lawIndex <- lawRename %>%
   mutate(twoPhysI = case_when(str_detect(twoPhys, "20 weeks")~2,
                               str_detect(twoPhys, "24 weeks")~1,
                               str_detect(twoPhys, "3rd trimester")~1,
                               str_detect(twoPhys, "Viability")~1),
          twoPhysI = replace_na(twoPhysI, 0)) %>%
          
                               
          
   mutate(gestLimI = case_when(str_detect(gestLim, "6 weeks")~4,
                               str_detect(gestLim,"20 weeks")~2, 
                               str_detect(gestLim, "24 weeks")~1, 
                               str_detect(gestLim, "3rd trimester")~1, 
                               str_detect(gestLim, "Viability")~1), 
          gestLimI = replace_na(gestLimI, 0)) %>%
  
   mutate(pbBanI = case_when(str_detect(pbBan, "X")~2,
                             str_detect(pbBan, "Postviability")~1),
          pbBanI = replace_na(pbBanI, 0)) %>%              
   
   mutate(pubFundAllI = case_when(str_detect(pubFundAll, "X")~0),
          pubFundAllI = replace_na(pubFundAllI, 1)) %>%
   
   mutate(pubFundLimI = case_when(str_detect(pubFundLim, "Life Only")~2,
                                  str_detect(pubFundLim, "X")~1),
          pubFundLimI = replace_na(pubFundLimI, 0)) %>%
                
   
   mutate(privLimI = case_when(str_detect(privLim, "X")~1),
          privLimI = replace_na(privLimI, 0)) %>%
   
   mutate(parentI = case_when(str_detect(parent, "Consent and Notice")~2,
                              str_detect(parent, "Consent")~2,
                              str_detect(parent, "Notice")~1),
          parentI = replace_na(parentI, 0)) %>%
   
   mutate(twoTripsI = case_when(str_detect(twoTrips, "X100")~3,
                                str_detect(twoTrips, "Xany")~2,
                                str_detect(twoTrips, "Xtele")~1,
                                str_detect(twoTrips, "X")~4), # X at end to include other entries starting with X
          twoTripsI = replace_na(twoTripsI, 0))
                                
                                 
```

```{r}
# Modify values for exceptions/special characters

#lawIndexM <- lawIndex %>%   
   #mutate(gestLimI = gestLimI+case_when(str_detect(gestLim, "[‡]")~0.5, #life endangerment only
                                        #str_detect(gestLim, "[€]")~0.25, #temporary block on 15 week ban
                                        #str_detect(gestLim, "[†]")~-0.25, #exception:rape or incest 
                                        #TRUE~0)
                              #+ case_when(str_detect(gestLim, "[*]")~-0.25,#exception:fetal abnormality
                                          #str_detect(gestLim, "[Ω]")~-0.25, #exception:patient's phys health          
                                       #TRUE~0)) %>% #add zero to keep gestLimI value
   
lawIndexM <- lawIndex %>%   
   mutate(gestLimI = gestLimI+case_when(str_detect(gestLim, "[‡]")~0.5, #life endangerment only
                                        TRUE~0)
          +case_when(str_detect(gestLim, "[€]")~0.25, #temporary block on 15 week ban
                                        TRUE~0)
          +case_when(str_detect(gestLim, "[†]")~-0.25, #exception:rape or incest
                                        TRUE~0)
          +case_when(str_detect(gestLim, "[*]")~-0.25, #exception:fetal abnormality
                                        TRUE~0)
          + case_when(str_detect(gestLim, "[Ω]")~-0.25, #exception:patient's phys health
                                       TRUE~0))  %>% # if no special characters, add zero to keep value
   
                                       
   mutate(pbBanI = pbBanI+case_when(str_detect(pbBan, "[▼]")~0.25,
                                    TRUE~0)) %>%  #law not in effect
   
   mutate(pubFundLimI = pubFundLimI+case_when(str_detect(pubFundLim, "[ϴ]")~0.25, #against court order
                                              str_detect(pubFundLim, "[*]")~-0.25, #exception:patient's phys health
                                              TRUE~0)
                                    + case_when(str_detect(pubFundLim, "[Ω]")~-0.25, #exception:fetal abnormality
                                       TRUE~0)) %>% 
   
   mutate(parentI = parentI+case_when(str_detect(parent, "[▼]")~0.25, #law not in effect
                                      str_detect(parent, "[ξ]")~-0.25, #may be waived
                                      str_detect(parent, "[β]")~-0.25, #under age 16
                                      TRUE~0)) %>% 

   mutate(twoTripsI = twoTripsI+case_when(str_detect(twoTrips, "[▼]")~0.25,
                                          TRUE~0))
   
```


```{r}
## save modified index with temporary adjustment values

write_csv(lawIndexM, here("code/data/tempLawIndex.csv"))
```

## Create Index

Group variables into categories
1. Gestational Limits
2. Funding
   - Public funding (all)
   - Public funding limited
   - Private insurance limited
3. Medical Regulations
   - Second Physician (if at)
   - Partial Birth Ban
4. Personal Restrictions
   - Parental involvement
   - In-person counseling req. two trips to clinic
   
```{r}
indexActive <- lawIndexM[-c(1:9)] # select only indexed variables, take out states b/c character 


# rescale columns of a dataframe to be 0 to 1 in R tidyverse

# define rescale function
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}


for(i in seq_along(indexActive)) { #iterate over columns
   indexActive[[i]] <- rescale01(indexActive[[i]])
}

row.names(indexActive) <- lawIndexM %>% pull(state)
```


```{r}
# add columns for each category

indexActive <- indexActive %>%
   mutate(funding= pubFundAllI + pubFundLimI + privLimI,
          medical= twoPhysI + pbBanI,
          personal= parentI + twoTripsI) %>%
   select(gestational = gestLimI, funding, medical, personal)

# rescale between 0 and 1

for(i in seq_along(indexActive)) { #iterate over columns
   indexActive[[i]] <- rescale01(indexActive[[i]])
}

# include state names 
indexActive <- indexActive %>% 
   mutate(state_name = pull(lawIndexM, state))

```

```{r}
# create summary column

indexFinal <- indexActive %>%
   filter(!state_name %in% c("AK", "HI", "DC")) %>% # remove states
   mutate(overall = gestational + funding + medical + personal) %>%
   arrange(desc(overall))
   
indexFinal %>%
   ggplot(aes(x= overall)) +
   geom_histogram() 

# summary stats
indexFinal %>%
   group_by(NULL) %>%
   summarise(mean= mean(overall),
             sd= sd(overall),
             count= n()
             )

# add state party

stateParty <- read_csv("data/stateParty.csv")

# combine data
indexFinal %>% 
   inner_join(stateParty, by= c("state_name" = "stateAB")) %>%
    group_by(stateCon) %>%
   summarise(mean= mean(overall),
             sd= sd(overall),
             count= n()
             )

# statistical significance? 
# Null hypothesis is no difference. 

(2.3151042 - 0.6104167)/(sqrt(0.6466908^2/20 + 0.5817762^2/14))	# denom = standard deviation

# 8.028269
```

